Probabilistic forecasting of hourly Emergency Department arrivals

### 2023-04-13 journal_submission/revised/latex_with_authors/paper_with_authors.tex
prepare a version to share in the journal

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
% TODO: Currently lineno needs to be loaded after amsmath because of conflict
% https://github.com/latex-lineno/lineno/issues/5
\usepackage{lineno} % add
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[top=25mm, left=25mm, right=25mm, bottom=25mm,headsep=10mm, footskip=12mm]{geometry}
\usepackage[]{natbib}
\bibliographystyle{plainnat}

@@

@@

Forecasting Emergency Department (ED) arrivals is critical for informing
staffing and scheduling decisions to meet the needs of patients.
Accurate ED demand forecasts contribute to a better decision-making
process regarding resources allocation and staffing. This is one of the
best ways to optimise resources utilisation and minimise related costs.
An accurate forecast of patient arrivals is crucial in ED services to
depict various courses of action that can result in massive savings in
terms of patient lives. Inability to match the staff with the demand
might result in patients overcrowding the system, which is a severe
problem that causes challenges for the patient flow
\citep{derlet2002overcrowding}. Also, it is related to increasing length of
stay \citep{muhammet2015forecasting}, low patient satisfaction, unexpected
return visits to services, increased health care costs, inaccuracy in
electronic medical records and other \citep{ROSTAMITABAR20221197}.

@@

There are few studies that look at forecasting hourly arrivals in ED and
other hospital services using historical time series data and/or
predictors such as patient characteristics, weather, holidays and public
events. These studies use multiple approaches, including Exponential
Smoothing \citep{SvetunkovAdam}, Autoregressive Integrated Moving Average
(ARIMA) \citep{hyndman2021forecasting}, Autoregressive Conditional
Heteroskedasticity (ARCH) \citep{bollerslev1994arch}, Vector Autoregressive
model \citep{lutkepohl2013vector}, TBATS \citep{de2011forecasting} and Artificial
Neural Networks \citep{hyndman2021forecasting}. \textcolor{blue}{We have identified a number of limitations and gaps in the area of ED forecasting that motivated us to undertake this study. These gaps and their importance are discussed below.} \textcolor{blue}{Most of these studies are limited to only predicting future arrivals as a point
forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals.} There are few studies that report uncertainty by presenting prediction intervals, but \textcolor{blue}{There is no study generating and evaluating the entire forecast distribution of arrivals. Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric, i.e. having more staff than needed is costly, but
having less staff than required may lead to worse outcomes for patients}. This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient \citep{wright2006reexamining}. Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently \citep{ramos2013probabilistic, ROSTAMITABAR20221197}. Furthermore, if the impact of under- and overstaffing can be quantified,
probabilistic forecasts allow \emph{optimal} decisions that balance the cost associated with under- and overstaffing to be calculated. Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task. \textcolor{blue}{Another drawback of existing studies is that the datasets used are relatively small (e.g. time period of 1-2 years), making it challenging to capture the inter-annual seasonality correctly and to report the forecast accuracy using robust approaches such as time series cross-validation. Such results might not be generalisable. Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals. Moreover, hourly ED time series may contain low volume values and zeros in some hours of the day, which brings additional challenges to traditional time series forecasting approaches. Finally, all previous publications referenced in this paper are not fully reproducible as underlying data, functions and code are not available.}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We produce probabilistic forecasts, in addition to the point
  estimation, quantifying uncertainties in future hospital admission,
  and comparing different forecasting methods using a suite of
  well-established evaluation metrics;
\item
  We develop an advanced dynamic model to forecast ED arrivals based
  on iETS \citep{Svetunkov2019a} and ETSX models with a modification for
  multiple frequencies, which produced highly-accurate point
  forecasts;
\item
  We develop a novel model to produce a probabilistic forecast of ED
  arrivals based on Generalised Additive Models for Location Scale and
  Shape, which accounts for i) the bounded and non-Gaussian
  distribution of arrivals, ii) multiple seasonalities, weather and
  holiday effects, and iii) variation in forecast uncertainty;
\item
  We benchmark the accuracy of our model against appropriate models
  used when multiple seasonality is present, i.e.~Prophet, TBATS,
  Poisson Regression, Exponential Smoothing State Space model (ETS)
  and the simple empirical distribution of the arrivals;
\item
  We provide data and code enabling reproduction and refinement of the
  proposed approach and benchmarks. The proposed approach could also
  be generalised to forecast hourly requirements for longer horizons
  and in other services \citep{al2021empirical}, such as inpatient and
  outpatient care services, the number of attended incidents in
  ambulance services, or call volumes in clinical desk services.
\end{enumerate}

@@

\textcolor{blue}{There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare. Areas such as call volume arrivals, ambulance demand and Emergency Department forecasting have received a significant attention. We refer interested readers to some extensive reviews of the relevant literature by} \citet{mingliterature2022}, \citet{gul2020review} and \citet{ibrahim2016modeling}. \textcolor{blue}{The time granularity considered in these studies spans from hourly to yearly across different parts of healthcare. However, given the focus of this study, we only discuss hourly ED forecasting.}

\citet{hertzum2017forecasting} used linear regression, ARIMA, and Naïve to
investigate whether accurate hourly accident and emergency department
patient arrivals and occupancy forecasts can be generated using calendar
variables. \citet{hertzum2017forecasting} study found that patient arrival
variation is larger across the hours of the day than across the days of
the week and the months of the year. In terms of the hour of the day,
patient arrivals peaked around noon. For days of the week, Monday is the
busiest day, while weekends are the quietest ones. July-August are the
months with the highest number of patient arrivals, while January and
February have the lowest numbers. They indicate that regression and
ARIMA models performed similarly in modelling patient arrivals, while
ARIMA outperformed regression models in modelling accident and emergency
department occupancy.

\citet{choudhury2020forecasting} used ARIMA, Holt-Winters, TBATS, and neural
network methods to forecast hourly accident and emergency department
arrivals. ARIMA model was selected as the best fit model. Authors
claimed that ARIMA provided high and acceptable hourly ED forecasting
accuracy, even outperforming TBATS. \citet{Cheng2021} developed an ARIMA model
for ED occupancy with a seasonal component and exogenous variables,
which outperformed a rolling-average benchmark. They also produce
prediction intervals, a form of the probabilistic forecast, which were
found to be well-calibrated, a necessary property for such forecasts.

\citet{morzuch2006forecasting} used the Unobserved Components Model (UCM), in
which each component of the time series is separately modelled as
stochastic. Double-seasonal exponential smoothing and standard
Holt-Winters were used to forecast ED arrival for a horizon of 168
hours. The hourly data collected from an ED in Pennsylvania showed no
trend and two seasonal cycles: a within-day and a within-week seasonal
cycles. The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly
seasonality of the time series.

\citet{mccarthy2008challenge} employed a Poisson log-linear regression model,
including independent variables such as temporal factors (e.g.,
hour-of-day, day-of-week, type-of-day, season, and calendar year),
patient characteristics (i.e., age, gender, insurance status, triage
level, mode of arrival, and ambulance diversion status) and climatic
factors (i.e., temperature and precipitation) to model patient demand
for ED services. The authors produced probabilistic predictions in the
form of \(50\%\) and \(90\%\) prediction intervals for the number of hourly
arrivals. Hourly data of ED arrivals in the 1-year study period was
modelled and analysed, and it was suggested that the model could be used
for forecasting. However, model evaluation was performed in-sample on
only one year of data, so it is unclear how this approach would perform
in a forecasting setting or compare to simpler approaches. However, the
length of the time series in this study was very short (only one year),
which did not allow for a rigorous out-of-sample evaluation.

\citet{schweigler2009forecasting} investigated whether time series methods
could accurately generate short-term forecasts of ED bed occupancy. A
year-long dataset of hourly ED bed occupancy was collected from three
facilities. The authors implemented an hourly historical average model,
SARIMA model, and sinusoidal model with autocorrelated error for each
facility. The historical average model was based on the mean occupancy
for each site, for each hour of the day, while the sinusoidal model was
based on four parameters: an AR term, a sine coefficient, a cosine
coefficient and an intercept. They evaluated the forecast accuracy of
four and twelve hours forecast horizons using RMSE. They found that both
SARIMA and the sinusoidal models outperformed the historical average
(for example, at site 2, the two models improved by \(33\%\) the 12-hour
forecasts generated by the historical average).

\citet{kim2014predicting} compared different univariate and multivariate time
series forecasting techniques to predict patient volume for a Hospital
Medicine programme. The study evaluated linear regression, exponential
smoothing, ARIMA, SARIMA, Generalized Autoregressive Conditional
Heteroskedasticity (GARCH) and Vector Autoregressive (VAR) models to
forecast for 4 and 24 hours ahead. They used Mean Absolute Percentage
Error (MAPE) to report the forecast accuracy. The authors found that the
ARIMA outperformed all the other models.

@@

\citet{asheim2019real} developed a Poisson time-series regression model with
continuous day-of-week and week-of-year effects to implement a real-time
system that could forecast ED arrivals on 1, 2, 3 hours ahead. Measuring
the accuracy using the MAPE, \citet{asheim2019real} noticed that significant
improvement happened when the time of notification was incorporated into
the model, especially in the one-hour horizon.

\citet{cheng2021forecasting} used one year of ED visits time series to evaluate
the Rolling Average, SARIMAX, ARIMA, VAR and Holt-Winter to forecast ED
occupancy up to 4-hours ahead. The forecast accuracy is evaluated using
Mean Squared Error (MSE), Mean Absolute Error (MAE) and MAPE for point
forecast and coverage for prediction intervals of \(80\%\) and \(95\%\).
They show that SARIMAX provides a more accurate forecast of hourly ED
occupancy.

According to the studies mentioned above, it can be said that they have
shown complications in forecasting hourly patient accident and emergency
department visits, and the application of forecasting hourly patient
visits is not well established. Some of the studies claimed that the
accuracy of forecasting models on hourly accident and emergency
department data is low \citep{boyle2012predicting, hertzum2017forecasting},
while others mentioned that the accuracy of ED hourly forecast is at an
acceptable level \citep{choudhury2020forecasting, mccarthy2008challenge, schweigler2009forecasting}.

@@

It is clear that the number of arrivals has a sub-daily structure,
similar to the one summarised by \citet{hertzum2017forecasting}. The ED
arrivals decrease between midnight and early morning and then increase
until the evening, decreasing after that again. It is also clear that ED
service gets systematically more visits on Mondays between 8 a.m. and 5
p.m. Moreover, the number of arrivals around midnight is slightly higher
for Saturdays and Sundays.

@@

Figure \ref{fig:seasonplot-dofw} \textcolor{blue}{ illustrates the daily subseries plot, with the x-axis representing the date and y-axis
the ED arrivals. Each individual plot illustrates how arrivals change over time for a each day
of week from Monday to Saturday. The blue line shows the average arrival for the given day.
It is clear that ED arrivals on Mondays are higher than on other days. This is followed by
Saturday. This indicates that there are significantly more arrivals on Mondays and Saturdays
compared to the rest of the week. This might be due to the closure of General Practitioners
outpatient clinics over the weekends.}

@@

Finally, Figure \ref{fig:date-plot} presents the time plot of daily
arrivals. Each point represents one day, and points are shape-coded by
day-of-week to show the weekly cycle. The figure shows more arrivals on
Mondays than on other weekdays, agreeing with the previous findings.
Moreover, we can observe a long-term trend line and the significant
effect of some holidays. We can see that arrivals near Christmas and New
Year's day are significantly lower than other days of the year.
Moreover, Figure \ref{fig:date-plot} allows us to identify the impact
of special events on arrivals. For instance, we see that arrivals are
significantly low for 01-03 of March 2018. These days correspond to the
Storm Emma with heavy snowfall that resulted in travel disruption, mass
power outages, and schools closed in the UK \citep{stormemma2018}.

@@

We start with one of the simplest forecasting approaches used in
practice - assuming that in the next few hours, everything will be the
same as in the similar hours of a similar day in the past. This is
called ``Naïve''. In our case, given that we need a distribution of
values, we will use a modified approach, where the
empirical distribution of the hourly arrival time series is used to
forecast the future arrival distribution \citep{la2021new}. We consider the
empirical distribution of all available historic data (Benchmark-1) and
the empirical distribution of the most recent year of historic data on a
rolling basis (Benchmark-2) to capture potential changes in behaviour
over time.

@@

Regression is one of the most popular forecasting methods that use
explanatory variables to predict a variable of interest (in our case,
the ED arrivals). The classical linear regression model is formulated as
\begin{equation}
  {y}_t = \mathbf{x}_t' \boldsymbol{\beta} + \epsilon_t ,
\label{eq:linearRegression}
\end{equation} where \(\mathbf{x}_t\) is the vector of explanatory
variables, \(\boldsymbol{\beta}\) is the vector of parameters,
\(\epsilon_t\) is the error term, which is typically assumed to follow
Normal distribution with zero mean and a fixed variance, and \(t\) is the
time index. However, in the context of healthcare and ED arrivals, the
assumption of Normality is unrealistic because the number of admitted
patients is an integer and non-negative. So the linear regression model
should be substituted by some other model. One of the models that is
frequently used in practice is the Poisson regression \citep[see, for example,][]{mccarthy2008challenge}, which can be summarised as \begin{equation}
  \begin{aligned}
    & {y}_t \sim \mathrm{Poisson} \left( \lambda_t \right) \\
    & \log \lambda_t = \mathbf{x}_t' \boldsymbol{\beta}
  \end{aligned} \quad .
\label{eq:PoissonRegression}
\end{equation}

@@

The variables \emph{1--3} allow to model the seasonal patterns on the
appropriate level of detail throughout the year, while \emph{4} covers the
changes in admittance due to calendar events. Finally, \emph{5} is needed to
capture the potential phenomenon of change in admittance after the
holiday (e.g.~people might try not to go to the hospital on Christmas
eve and thus go the next day). This model assumes that all these effects
are deterministic and do not change over time. Still, the exponentiation
in \eqref{eq:PoissonRegression} introduces an interaction effect between
dummy variables so that the 3 pm on Monday in January will be different
from 3 pm on Monday in July, although the parameters for the hour of day
and day of the week are fixed and do not change over time. We use the
\texttt{alm()} function from the \texttt{greybox} package \citep{Svetunkov2021Greybox} for
R \citep{RTeam2021} for the experiments and denote this model as ``Poisson
Regression''.

@@

\citet{Hyndman2008b} developed a state space approach for exponential smoothing
models. The model can have a set of elements, including different types
of Error, Trend and Seasonal components (thus \emph{ETS}). Given the
popularity of the ETS model in the forecasting community, we decided to
include the basic ETS(A,N,A) model with the seasonal component with a
frequency of 24 (hour of the day) as a benchmark. This was done using
the \texttt{adam()} function from the \texttt{smooth} package \citep{Svetunkov2021Smooth}
for R and denoted as \emph{ETS}. This model does not capture the day of week
or week of year effects, does not include explanatory variables, but its
seasonal component and level change over time. This model is included as
a benchmark, only to see how the other models perform compared to it.

@@

Prophet is a forecasting procedure created by Facebook
\citep{taylor2018forecasting} that accounts for multiple seasonality,
piecewise trend and holiday effects. Prophet is robust to missing data
and shifts in the trend and typically handles outliers well. Prophet
works well on daily data seen in Facebook. It is robust and automated,
making it easy to learn for beginners. The implementation may be less
flexible than other methods. The model itself relies on the Multiple
Source of Error state space model, initially proposed by \citet{Kalman1960}.
The model is incorporated using the corresponding implementation of the
Fable package in R. We use the \texttt{prophet()} function from the \texttt{fable}
package \citep{fable2020}. Note that the input data is assigned with an
hourly and daily seasonality. \textcolor{blue}{This method has been adopted in some healthcare service providers in the United Kingdom to produce forecasts, therefore we have included it as one of our
benchmarks.}

@@

\textcolor{blue}{
The dataset used in this study does not include the period of the COVID-19 pandemic.
During COVID-19, the dynamics of ED arrivals has changed substantially. This means that
any forecasting model used for ED arrivals forecasting during that period would need to be
modified to reflect those changes for that specific period. One of the simplest modifications
would be to include a set of dummy variables, capturing different stages of the pandemic.
However, this is outside of the scope of this paper and can be considered as a direction for
future research.}

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
% TODO: Currently lineno needs to be loaded after amsmath because of conflict
% https://github.com/latex-lineno/lineno/issues/5
\usepackage{lineno} % add
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[top=25mm, left=25mm, right=25mm, bottom=25mm,headsep=10mm, footskip=12mm]{geometry}

@@

% Pandoc citation processing
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
% for Pandoc 2.8 to 2.10.1
\newenvironment{cslreferences}%
  {}%
  {\par}
% For Pandoc 2.11+
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

@@

Forecasting Emergency Department (ED) arrivals is critical for informing
staffing and scheduling decisions to meet the needs of patients.
Accurate ED demand forecasts contribute to a better decision-making
process regarding resources allocation and staffing. This is one of the
best ways to optimise resources utilisation and minimise related costs.
An accurate forecast of patient arrivals is crucial in ED services to
depict various courses of action that can result in massive savings in
terms of patient lives. Inability to match the staff with the demand
might result in patients overcrowding the system, which is a severe
problem that causes challenges for the patient flow
(Derlet, 2002). Also, it is related to increasing length of
stay (Muhammet and Guneri, 2015), low patient satisfaction, unexpected
return visits to services, increased health care costs, inaccuracy in
electronic medical records (Rostami-Tabar and Ziel, 2022).

@@

There are few studies that look at forecasting hourly arrivals in ED and
other hospital services using historical time series data and/or
predictors such as patient characteristics, weather, holidays and public
events. These studies use multiple approaches, including Exponential
Smoothing (Svetunkov, 2021a), Autoregressive Integrated Moving Average
(ARIMA) (Hyndman and Athanasopoulos, 2021), Autoregressive Conditional
Heteroskedasticity (ARCH) (Bollerslev et al., 1994), Vector Autoregressive
model (Lütkepohl, 2013), TBATS (De Livera et al., 2011) and Artificial
Neural Networks (Hyndman and Athanasopoulos, 2021).We have identified a number of limitations and gaps in the area of ED forecasting that motivated us to undertake this study. These gaps and their importance are discussed below. Most of these studies are limited to only predicting future arrivals as a point
forecast (a single number), which does not quantify any uncertainty associated with the number of future arrivals. There are few studies that report uncertainty by presenting prediction intervals, but there is no study generating and evaluating the entire forecast distribution of arrivals. Reporting the uncertainty via the forecast distribution is potentially valuable in this setting and has practical implications for those managing Emergency Departments because the consequences of inadequate staffing are asymmetric, i.e. having more staff than needed is costly, but
having less staff than required may lead to worse outcomes for patients. This asymmetry arises because it is preferable to incur a small opportunity cost associated with utilised staff rather than compromise service levels if staff levels are insufficient (Wright et al., 2006). Probabilistic forecasts inform decision-makers about exposure to these risks and potentially enable those risks to be managed more efficiently (Ramos et al., 2013; Rostami-Tabar and Ziel, 2022). Furthermore, if the impact of under- and overstaffing can be quantified,
probabilistic forecasts allow \emph{optimal} decisions that balance the cost associated with under- and overstaffing to be calculated. Therefore, in this paper, in addition to generating point forecasts, we also produce and evaluate density forecasts of hourly ED arrivals, comparing several methods for this task. Another drawback of existing studies is that the datasets used are relatively small (e.g. time period of 1-2 years), making it challenging to capture the inter-annual seasonality correctly and to report the forecast accuracy using robust approaches such as time series cross-validation. Such results might not be generalisable. Additionally, most of the forecasting methods used in these publications do not consider the full extent of the multiple seasonality of hourly ED arrivals. Moreover, hourly ED time series may contain low volume values and zeros in some hours of the day, which brings additional challenges to traditional time series forecasting approaches. Finally, all previous publications referenced in this paper are not fully reproducible as underlying data, functions and code are not available.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  We produce probabilistic forecasts, in addition to the point
  estimation, quantifying uncertainties in future hospital admission,
  and comparing different forecasting methods using a suite of
  well-established evaluation metrics;
\item
  We develop an advanced dynamic model to forecast ED arrivals based
  on iETS (Svetunkov and Boylan, 2019) and ETSX models with a modification for
  multiple frequencies, which produced highly-accurate point
  forecasts;
\item
  We develop a novel model to produce a probabilistic forecast of ED
  arrivals based on Generalised Additive Models for Location Scale and
  Shape, which accounts for i) the bounded and non-Gaussian
  distribution of arrivals, ii) multiple seasonalities, weather and
  holiday effects, and iii) variation in forecast uncertainty;
\item
  We benchmark the accuracy of our model against appropriate models
  used when multiple seasonality is present, i.e.~Prophet, TBATS,
  Poisson Regression, Exponential Smoothing State Space model (ETS)
  and the simple empirical distribution of the arrivals;
\item
  We provide data and code enabling reproduction and refinement of the
  proposed approach and benchmarks. The proposed approach could also
  be generalised to forecast hourly requirements for longer horizons
  and in other services (Al-Azzani et al., 2021), such as inpatient and
  outpatient care services, the number of attended incidents in
  ambulance services, or call volumes in clinical desk services.
\end{enumerate}

@@

There is a substantial number of studies that employ models to forecast admissions and arrivals to inform planning and decision making in the healthcare. Areas such as call volume arrivals, ambulance demand and Emergency Department forecasting have received a significant attention. We refer interested readers to some extensive reviews of the relevant literature by Shi et al. (2022), Gul and Celik (2020) and Ibrahim et al. (2016). The time granularity considered in these studies spans from hourly to yearly across different parts of healthcare. However, given the focus of this study, we only discuss hourly ED forecasting.

Hertzum (2017) used linear regression, ARIMA, and Naïve to
investigate whether accurate hourly accident and emergency department
patient arrivals and occupancy forecasts can be generated using calendar
variables. Hertzum (2017) study found that patient arrival
variation is larger across the hours of the day than across the days of
the week and the months of the year. In terms of the hour of the day,
patient arrivals peaked around noon. For days of the week, Monday is the
busiest day, while weekends are the quietest ones. July-August are the
months with the highest number of patient arrivals, while January and
February have the lowest numbers. They indicate that regression and
ARIMA models performed similarly in modelling patient arrivals, while
ARIMA outperformed regression models in modelling accident and emergency
department occupancy.

Choudhury and Urena (2020) used ARIMA, Holt-Winters, TBATS, and neural
network methods to forecast hourly accident and emergency department
arrivals. ARIMA model was selected as the best fit model. Authors
claimed that ARIMA provided high and acceptable hourly ED forecasting
accuracy, even outperforming TBATS. Cheng et al. (2021a) developed an ARIMA model
for ED occupancy with a seasonal component and exogenous variables,
which outperformed a rolling-average benchmark. They also produce
prediction intervals, a form of the probabilistic forecast, which were
found to be well-calibrated, a necessary property for such forecasts.

Morzuch and Allen (2006) used the Unobserved Components Model (UCM), in
which each component of the time series is separately modelled as
stochastic. Double-seasonal exponential smoothing and standard
Holt-Winters were used to forecast ED arrival for a horizon of 168
hours. The hourly data collected from an ED in Pennsylvania showed no
trend and two seasonal cycles: a within-day and a within-week seasonal
cycles. The double seasonal model recorded lower RMSEs for all the 168-hour horizons, which was expected due to the strong hourly
seasonality of the time series.

McCarthy et al. (2008) employed a Poisson log-linear regression model,
including independent variables such as temporal factors (e.g.,
hour-of-day, day-of-week, type-of-day, season, and calendar year),
patient characteristics (i.e., age, gender, insurance status, triage
level, mode of arrival, and ambulance diversion status) and climatic
factors (i.e., temperature and precipitation) to model patient demand
for ED services. The authors produced probabilistic predictions in the
form of \(50\%\) and \(90\%\) prediction intervals for the number of hourly
arrivals. Hourly data of ED arrivals in the 1-year study period was
modelled and analysed, and it was suggested that the model could be used
for forecasting. However, model evaluation was performed in-sample on
only one year of data, so it is unclear how this approach would perform
in a forecasting setting or compare to simpler approaches. However, the
length of the time series in this study was very short (only one year),
which did not allow for a rigorous out-of-sample evaluation.

Schweigler et al. (2009) investigated whether time series methods
could accurately generate short-term forecasts of ED bed occupancy. A
year-long dataset of hourly ED bed occupancy was collected from three
facilities. The authors implemented an hourly historical average model,
SARIMA model, and sinusoidal model with autocorrelated error for each
facility. The historical average model was based on the mean occupancy
for each site, for each hour of the day, while the sinusoidal model was
based on four parameters: an AR term, a sine coefficient, a cosine
coefficient and an intercept. They evaluated the forecast accuracy of
four and twelve hours forecast horizons using RMSE. They found that both
SARIMA and the sinusoidal models outperformed the historical average
(for example, at site 2, the two models improved by \(33\%\) the 12-hour
forecasts generated by the historical average).

Kim et al. (2014) compared different univariate and multivariate time
series forecasting techniques to predict patient volume for a Hospital
Medicine programme. The study evaluated linear regression, exponential
smoothing, ARIMA, SARIMA, Generalized Autoregressive Conditional
Heteroskedasticity (GARCH) and Vector Autoregressive (VAR) models to
forecast for 4 and 24 hours ahead. They used Mean Absolute Percentage
Error (MAPE) to report the forecast accuracy. The authors found that the
ARIMA outperformed all the other models.

@@

Asheim et al. (2019) developed a Poisson time-series regression model with
continuous day-of-week and week-of-year effects to implement a real-time
system that could forecast ED arrivals on 1, 2, 3 hours ahead. Measuring
the accuracy using the MAPE, Asheim et al. (2019) noticed that significant
improvement happened when the time of notification was incorporated into
the model, especially in the one-hour horizon.

Cheng et al. (2021b) used one year of ED visits time series to evaluate
the Rolling Average, SARIMAX, ARIMA, VAR and Holt-Winter to forecast ED
occupancy up to 4-hours ahead. The forecast accuracy is evaluated using
Mean Squared Error (MSE), Mean Absolute Error (MAE) and MAPE for point
forecast and coverage for prediction intervals of \(80\%\) and \(95\%\).
They show that SARIMAX provides a more accurate forecast of hourly ED
occupancy.

According to the studies mentioned above, it can be said that they have
shown complications in forecasting hourly patient accident and emergency
department visits, and the application of forecasting hourly patient
visits is not well established. Some of the studies claimed that the
accuracy of forecasting models on hourly accident and emergency
department data is low (Boyle et al., 2012; Hertzum, 2017),
while others mentioned that the accuracy of ED hourly forecast is at an
acceptable level (Choudhury and Urena, 2020; McCarthy et al., 2008; Schweigler et al., 2009).

@@

It is clear that the number of arrivals has a sub-daily structure,
similar to the one summarised by Hertzum (2017). The ED
arrivals decrease between midnight and early morning and then increase
until the evening, decreasing after that again. It is also clear that ED
service gets systematically more visits on Mondays between 8 a.m. and 5
p.m. Moreover, the number of arrivals around midnight is slightly higher
for Saturdays and Sundays.

@@

Figure \ref{fig:seasonplot-dofw} illustrates the daily subseries plot, with the x-axis representing the date and y-axis
the ED arrivals. Each individual plot illustrates how arrivals change over time for a each day
of week from Monday to Saturday. The blue line shows the average arrival for the given day.
It is clear that ED arrivals on Mondays are higher than on other days. This is followed by
Saturday. This indicates that there are significantly more arrivals on Mondays and Saturdays
compared to the rest of the week. This might be due to the closure of General Practitioners
outpatient clinics over the weekends.

@@

Finally, Figure \ref{fig:date-plot} presents the time plot of daily
arrivals. Each point represents one day, and points are shape-coded by
day-of-week to show the weekly cycle. The figure shows more arrivals on
Mondays than on other weekdays, agreeing with the previous findings.
Moreover, we can observe a long-term trend line and the significant
effect of some holidays. We can see that arrivals near Christmas and New
Year's day are significantly lower than other days of the year.
Moreover, Figure \ref{fig:date-plot} allows us to identify the impact
of special events on arrivals. For instance, we see that arrivals are
significantly low for 01-03 of March 2018. These days correspond to the
Storm Emma with heavy snowfall that resulted in travel disruption, mass
power outages, and schools closed in the UK (Wales, n.d.).

@@

We start with one of the simplest forecasting approaches used in
practice - assuming that in the next few hours, everything will be the
same as in the similar hours of a similar day in the past. This is
called ``Naïve''. In our case, given that we need a distribution of
values, we will use a modified approach, where the
empirical distribution of the hourly arrival time series is used to
forecast the future arrival distribution (La Salle et al., 2021). We consider the
empirical distribution of all available historic data (Benchmark-1) and
the empirical distribution of the most recent year of historic data on a
rolling basis (Benchmark-2) to capture potential changes in behaviour
over time.

@@

Regression is one of the most popular forecasting methods that use
explanatory variables to predict a variable of interest (in our case,
the ED arrivals). The classical linear regression model is formulated as
\begin{equation}
  {y}_t = \mathbf{x}_t' \boldsymbol{\beta} + \epsilon_t ,
\label{eq:linearRegression}
\end{equation} where \(\mathbf{x}_t\) is the vector of explanatory
variables, \(\boldsymbol{\beta}\) is the vector of parameters,
\(\epsilon_t\) is the error term, which is typically assumed to follow
Normal distribution with zero mean and a fixed variance, and \(t\) is the
time index. However, in the context of healthcare and ED arrivals, the
assumption of Normality is unrealistic because the number of admitted
patients is an integer and non-negative. So the linear regression model
should be substituted by some other model. One of the models that is
frequently used in practice is the Poisson regression (see, for example, McCarthy et al., 2008), which can be summarised as \begin{equation}
  \begin{aligned}
    & {y}_t \sim \mathrm{Poisson} \left( \lambda_t \right) \\
    & \log \lambda_t = \mathbf{x}_t' \boldsymbol{\beta}
  \end{aligned} \quad .
\label{eq:PoissonRegression}
\end{equation}

@@

The variables \emph{1--3} allow to model the seasonal patterns on the
appropriate level of detail throughout the year, while \emph{4} covers the
changes in admittance due to calendar events. Finally, \emph{5} is needed to
capture the potential phenomenon of change in admittance after the
holiday (e.g.~people might try not to go to the hospital on Christmas
eve and thus go the next day). This model assumes that all these effects
are deterministic and do not change over time. Still, the exponentiation
in \eqref{eq:PoissonRegression} introduces an interaction effect between
dummy variables so that the 3 pm on Monday in January will be different
from 3 pm on Monday in July, although the parameters for the hour of day
and day of the week are fixed and do not change over time. We use the
\texttt{alm()} function from the \texttt{greybox} package (Svetunkov, 2021b) for
R (Team, 2021) for the experiments and denote this model as ``Poisson
Regression''.

@@

Hyndman et al. (2008) developed a state space approach for exponential smoothing
models. The model can have a set of elements, including different types
of Error, Trend and Seasonal components (thus \emph{ETS}). Given the
popularity of the ETS model in the forecasting community, we decided to
include the basic ETS(A,N,A) model with the seasonal component with a
frequency of 24 (hour of the day) as a benchmark. This was done using
the \texttt{adam()} function from the \texttt{smooth} package (Svetunkov, 2021c)
for R and denoted as \emph{ETS}. This model does not capture the day of week
or week of year effects, does not include explanatory variables, but its
seasonal component and level change over time. This model is included as
a benchmark, only to see how the other models perform compared to it.

@@

Prophet is a forecasting procedure created by Facebook
(Taylor and Letham, 2018) that accounts for multiple seasonality,
piecewise trend and holiday effects. Prophet is robust to missing data
and shifts in the trend and typically handles outliers well. Prophet
works well on daily data seen in Facebook. It is robust and automated,
making it easy to learn for beginners. The implementation may be less
flexible than other methods. The model itself relies on the Multiple
Source of Error state space model, initially proposed by Kalman (1960).
The model is incorporated using the corresponding implementation of the
Fable package in R. We use the \texttt{prophet()} function from the \texttt{fable}
package (O'Hara-Wild et al., 2020). Note that the input data is assigned with an
hourly and daily seasonality. This method has been adopted in some healthcare service providers in the United Kingdom to produce forecasts, therefore we have included it as one of our
benchmarks.

@@

The dataset used in this study does not include the period of the COVID-19 pandemic.
During COVID-19, the dynamics of ED arrivals has changed substantially. This means that
any forecasting model used for ED arrivals forecasting during that period would need to be
modified to reflect those changes for that specific period. One of the simplest modifications
would be to include a set of dummy variables, capturing different stages of the pandemic.
However, this is outside of the scope of this paper and can be considered as a direction for
future research.
